\texttt{JAXbind}: Easy bindings to JAX

### 2024-03-12 paper/paper.tex

\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={jax\_op: Use any function in JAX},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\newenvironment{Shaded}{}{}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{#1}}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.49,0.56,0.16}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.53,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.56,0.13,0.00}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.73,0.13,0.13}{\textit{#1}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.02,0.16,0.49}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.40,0.40,0.40}{#1}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.74,0.48,0.00}{#1}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.73,0.40,0.53}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.10,0.09,0.49}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}

@@

\title{\texttt{jax\_op}: Use any function in JAX}

@@

JAX is widely used in machine learning and scientific computing.
Scientific computing relies on existing high-performance code which we
would ideally like to use in JAX. Reimplementing the existing code in
JAX is often impractical and the existing interface in JAX for
connecting custom code requires deep knowledge of JAX and its C++
backend. The aim of \texttt{jax\_op} is to drastically lower the burden
of connecting custom functions implemented in other programming
languages to JAX. Specifically, \texttt{jax\_op} provides an easy-to-use
Python interface for defining custom, so-called JAX primitives
supporting any JAX transformations.

@@

In this paper, we present \texttt{jax\_op}, a package for bridging any
function to JAX without in-depth knowledge of JAX's transformation
system. The interface is accessible from Python without requiring any
development in C++. The package is able to register any function, its
partial derivatives and their transpose functions as a JAX native call,
a so-called primitive. Derivatives, compilation rules, and batching
rules are automatically registered with JAX.

We believe \texttt{jax\_op} to be highly useful in scientific computing.
We intend to use this package to connect the Hartley transform and the
spherical harmonic transform from ducc (Reinecke, 2024) to NIFTy
(Edenhofer et al., 2024) as well as the radio interferometry response
from ducc with \texttt{resolve} (Arras et al., 2024) for radio
astronomy. Furthermore, we intend to connect the non-uniform FFT from
ducc with JAX for applications in strong-lensing astrophysics. We
envision many further applications within and outside of astrophysics.

@@

Automatic differentiation is a core feature of JAX and often one of the
main reasons for using it. Thus, it is essential that custom functions
registered with JAX support automatic differentiation. In the following,
we will outline which functions our package respectively JAX requires to
enable automatic differentiation. For simplicity, we assume that we want
to connect the nonlinear function \(f(x_1,x_2) = x_1x_2^2\) to JAX. The
\texttt{jax\_op} package expects the Python function for \(f\) to take
three positional arguments. The first argument, \texttt{out}, is a
\texttt{tuple} into which the function results are written. The second
argument is also a \texttt{tuple} containing the input to the function,
in our case, \(x_1\) and \(x_2\). Via \texttt{kwargs\_dump}, potential
keyword arguments given to the later registered Jax primitive can be
forwarded to \texttt{f} in serialized form.

\begin{Shaded}
\begin{Highlighting}[]
\ImportTok{import}\NormalTok{ jax\_linop }\CommentTok{\#FIXME}

\KeywordTok{def}\NormalTok{ f(out, args, kwargs\_dump):}
\NormalTok{    kwargs }\OperatorTok{=}\NormalTok{ jax\_linop.load\_kwargs(kwargs\_dump)}
\NormalTok{    x1, x2 }\OperatorTok{=}\NormalTok{ args}
\NormalTok{    out[}\DecValTok{0}\NormalTok{][()] }\OperatorTok{=}\NormalTok{ x1 }\OperatorTok{*}\NormalTok{ x2}\OperatorTok{**}\DecValTok{2}
\end{Highlighting}
\end{Shaded}

JAX's automatic differentiation engine can compute the Jacobian-vector
product \texttt{jvp} and vector-Jacobian product \texttt{vjp} of JAX
primitives. The Jacobian-vector product in JAX is a function applying
the Jacobian of \(f\) at a position \(x\) to a tangent vector. In
mathematical nomenclature this operation is called the pushforward of
\(f\) and can be denoted as \(\partial f(x): T_x X \mapsto T_{f(x)} Y\),
with \(T_x X\) and \(T_{f(x)} Y\) being the tangent spaces of \(X\) and
\(Y\) at the positions \(x\) and \(f(x)\). As the implementation of
\(f\) is not JAX native, JAX cannot automatically compute the
\texttt{jvp}. Instead, an implementation of the pushforward has to be
provided, which \texttt{jax\_op} will register as the \texttt{jvp} of
the JAX primitive of \(f\). For our example, this
Jacobian-vector-product function is given by
\(\partial f(x_1,x_2)(dx_1,dx_2) = x_2^2dx_1 + 2x_1x_2dx_2\).

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{def}\NormalTok{ f\_jvp(out, args, kwargs\_dump):}
\NormalTok{    kwargs }\OperatorTok{=}\NormalTok{ jax\_linop.load\_kwargs(kwargs\_dump)}
\NormalTok{    x1, x2, dx1, dx2 }\OperatorTok{=}\NormalTok{ args}
\NormalTok{    out[}\DecValTok{0}\NormalTok{][()] }\OperatorTok{=}\NormalTok{ x2}\OperatorTok{**}\DecValTok{2} \OperatorTok{*}\NormalTok{ dx1 }\OperatorTok{+} \DecValTok{2} \OperatorTok{*}\NormalTok{ x1 }\OperatorTok{*}\NormalTok{ x2 }\OperatorTok{*}\NormalTok{ dx2}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{def}\NormalTok{ f\_vjp(out, args, kwargs\_dump):}
\NormalTok{    kwargs }\OperatorTok{=}\NormalTok{ jax\_linop.load\_kwargs(kwargs\_dump)}
\NormalTok{    x1, x2, dy }\OperatorTok{=}\NormalTok{ args}
\NormalTok{    out[}\DecValTok{0}\NormalTok{][()] }\OperatorTok{=}\NormalTok{ x2}\OperatorTok{**}\DecValTok{2} \OperatorTok{*}\NormalTok{ dy}
\NormalTok{    out[}\DecValTok{1}\NormalTok{][()] }\OperatorTok{=} \DecValTok{2} \OperatorTok{*}\NormalTok{ x1 }\OperatorTok{*}\NormalTok{ x2 }\OperatorTok{*}\NormalTok{ dy}
\end{Highlighting}
\end{Shaded}

@@

We have now defined all ingredients necessary to register a JAX
primitive for our function \(f\) using the \texttt{jax\_op} package.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{f\_jax }\OperatorTok{=}\NormalTok{ jax\_linop.get\_nonlinear\_call(}
\NormalTok{    f, (f\_jvp, f\_vjp), f\_abstract, f\_abstract\_T}
\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\texttt{f\_jax} is a JAX primitive registered via the \texttt{jax\_op}
package supporting all JAX transformations. We can now compute the
\texttt{jvp} and \texttt{vjp} of the new JAX primitive and even
jit-compile and batch it.

@@

JAX supports higher order derivatives and can differentiate a
\texttt{jvp} or \texttt{vjp} with respect to the position at which the
Jacobian was taken. Similar to first derivatives, JAX can not
automatically compute higher derivatives of a general function \(f\)
that is not natively implemented in JAX. Higher order derivatives would
again need to be provided by the user. For many algorithms, first
derivatives are sufficient, and higher order derivatives are often not
implemented by the high-performance codes. Therefore, the current
interface of \texttt{jax\_op} is, for simplicity, restricted to first
derivatives. In the future, the interface could be easily expanded if
specific use cases require higher order derivatives.

In scientific computing, linear functions such as, e.g., spherical
harmonic transforms are widespread. If the function \(f\) is linear,
differentiation becomes trivial. Specifically for a linear function
\(f\), the pushforward respectively the \texttt{jvp} of \(f\) is
identical to \(f\) itself and independent of the position at which it is
computed. Expressed in formulas, \(\partial f(x)(dx) = f(dx)\) if \(f\)
is linear in \(x\). Analogously, the pullback respectively the
\texttt{vjp} becomes independent of the initial position and is given by
the linear transpose of \(f\), thus
\((\partial f(x))^{T}(dy) = f^T(dy)\). Also, all higher order
derivatives can be expressed in terms of \(f\) and its transpose. To
make use of these simplifications, \texttt{jax\_op} provides a special
interface for linear functions, supporting higher order derivatives,
only requiring an implementation of the function and its transpose.

@@

Currently, \texttt{jax\_op} only has CPU but no GPU support. With some
expertise on Python bindings for GPU kernels adding GPU support should
be fairly simple. We especially want to highlight that the interfacing
with the JAX automatic differentiation engine is identical for CPU and
GPU.

>>>>

\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={JAXbind: Easy bindings to JAX},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\newenvironment{Shaded}{}{}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{#1}}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.49,0.56,0.16}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.53,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.56,0.13,0.00}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.73,0.13,0.13}{\textit{#1}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{1.00,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.25,0.63,0.44}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.02,0.16,0.49}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.40,0.40,0.40}{#1}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.44,0.13}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.74,0.48,0.00}{#1}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.73,0.40,0.53}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.10,0.09,0.49}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.25,0.44,0.63}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.38,0.63,0.69}{\textbf{\textit{#1}}}}

@@

\title{\texttt{JAXbind}: Easy bindings to JAX}

@@

JAX is widely used in machine learning and scientific computing.
Scientific computing relies on existing high-performance code which we
would ideally like to use in JAX. Reimplementing the existing code in
JAX is often impractical and the existing interface in JAX for
connecting custom code requires deep knowledge of JAX and its C++
backend. The aim of \texttt{JAXbind} is to drastically lower the burden
of connecting custom functions implemented in other programming
languages to JAX. Specifically, \texttt{JAXbind} provides an easy-to-use
Python interface for defining custom, so-called JAX primitives
supporting any JAX transformations.

@@

In this paper, we present \texttt{JAXbind}, a package for bridging any
function to JAX without in-depth knowledge of JAX's transformation
system. The interface is accessible from Python without requiring any
development in C++. The package is able to register any function, its
partial derivatives and their transpose functions as a JAX native call,
a so-called primitive. Derivatives, compilation rules, and batching
rules are automatically registered with JAX.

We believe \texttt{JAXbind} to be highly useful in scientific computing.
We intend to use this package to connect the Hartley transform and the
spherical harmonic transform from ducc (Reinecke, 2024) to NIFTy
(Edenhofer et al., 2024) as well as the radio interferometry response
from ducc with \texttt{resolve} (Arras et al., 2024) for radio
astronomy. Furthermore, we intend to connect the non-uniform FFT from
ducc with JAX for applications in strong-lensing astrophysics. We
envision many further applications within and outside of astrophysics.

@@

Automatic differentiation is a core feature of JAX and often one of the
main reasons for using it. Thus, it is essential that custom functions
registered with JAX support automatic differentiation. In the following,
we will outline which functions our package respectively JAX requires to
enable automatic differentiation. For simplicity, we assume that we want
to connect the nonlinear function \(f(x_1,x_2) = x_1x_2^2\) to JAX. The
\texttt{JAXbind} package expects the Python function for \(f\) to take
three positional arguments. The first argument, \texttt{out}, is a
\texttt{tuple} into which the function results are written. The second
argument is also a \texttt{tuple} containing the input to the function,
in our case, \(x_1\) and \(x_2\). Via \texttt{kwargs\_dump}, potential
keyword arguments given to the later registered Jax primitive can be
forwarded to \texttt{f} in serialized form.

\begin{Shaded}
\begin{Highlighting}[]
\ImportTok{import}\NormalTok{ jaxbind }

\KeywordTok{def}\NormalTok{ f(out, args, kwargs\_dump):}
\NormalTok{    kwargs }\OperatorTok{=}\NormalTok{ jaxbind.load\_kwargs(kwargs\_dump)}
\NormalTok{    x1, x2 }\OperatorTok{=}\NormalTok{ args}
\NormalTok{    out[}\DecValTok{0}\NormalTok{][()] }\OperatorTok{=}\NormalTok{ x1 }\OperatorTok{*}\NormalTok{ x2}\OperatorTok{**}\DecValTok{2}
\end{Highlighting}
\end{Shaded}

JAX's automatic differentiation engine can compute the Jacobian-vector
product \texttt{jvp} and vector-Jacobian product \texttt{vjp} of JAX
primitives. The Jacobian-vector product in JAX is a function applying
the Jacobian of \(f\) at a position \(x\) to a tangent vector. In
mathematical nomenclature this operation is called the pushforward of
\(f\) and can be denoted as \(\partial f(x): T_x X \mapsto T_{f(x)} Y\),
with \(T_x X\) and \(T_{f(x)} Y\) being the tangent spaces of \(X\) and
\(Y\) at the positions \(x\) and \(f(x)\). As the implementation of
\(f\) is not JAX native, JAX cannot automatically compute the
\texttt{jvp}. Instead, an implementation of the pushforward has to be
provided, which \texttt{JAXbind} will register as the \texttt{jvp} of
the JAX primitive of \(f\). For our example, this
Jacobian-vector-product function is given by
\(\partial f(x_1,x_2)(dx_1,dx_2) = x_2^2dx_1 + 2x_1x_2dx_2\).

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{def}\NormalTok{ f\_jvp(out, args, kwargs\_dump):}
\NormalTok{    kwargs }\OperatorTok{=}\NormalTok{ jaxbind.load\_kwargs(kwargs\_dump)}
\NormalTok{    x1, x2, dx1, dx2 }\OperatorTok{=}\NormalTok{ args}
\NormalTok{    out[}\DecValTok{0}\NormalTok{][()] }\OperatorTok{=}\NormalTok{ x2}\OperatorTok{**}\DecValTok{2} \OperatorTok{*}\NormalTok{ dx1 }\OperatorTok{+} \DecValTok{2} \OperatorTok{*}\NormalTok{ x1 }\OperatorTok{*}\NormalTok{ x2 }\OperatorTok{*}\NormalTok{ dx2}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{def}\NormalTok{ f\_vjp(out, args, kwargs\_dump):}
\NormalTok{    kwargs }\OperatorTok{=}\NormalTok{ jaxbind.load\_kwargs(kwargs\_dump)}
\NormalTok{    x1, x2, dy }\OperatorTok{=}\NormalTok{ args}
\NormalTok{    out[}\DecValTok{0}\NormalTok{][()] }\OperatorTok{=}\NormalTok{ x2}\OperatorTok{**}\DecValTok{2} \OperatorTok{*}\NormalTok{ dy}
\NormalTok{    out[}\DecValTok{1}\NormalTok{][()] }\OperatorTok{=} \DecValTok{2} \OperatorTok{*}\NormalTok{ x1 }\OperatorTok{*}\NormalTok{ x2 }\OperatorTok{*}\NormalTok{ dy}
\end{Highlighting}
\end{Shaded}

@@

We have now defined all ingredients necessary to register a JAX
primitive for our function \(f\) using the \texttt{JAXbind} package.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{f\_jax }\OperatorTok{=}\NormalTok{ jaxbind.get\_nonlinear\_call(}
\NormalTok{    f, (f\_jvp, f\_vjp), f\_abstract, f\_abstract\_T}
\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\texttt{f\_jax} is a JAX primitive registered via the \texttt{JAXbind}
package supporting all JAX transformations. We can now compute the
\texttt{jvp} and \texttt{vjp} of the new JAX primitive and even
jit-compile and batch it.

@@

JAX supports higher order derivatives and can differentiate a
\texttt{jvp} or \texttt{vjp} with respect to the position at which the
Jacobian was taken. Similar to first derivatives, JAX can not
automatically compute higher derivatives of a general function \(f\)
that is not natively implemented in JAX. Higher order derivatives would
again need to be provided by the user. For many algorithms, first
derivatives are sufficient, and higher order derivatives are often not
implemented by the high-performance codes. Therefore, the current
interface of \texttt{JAXbind} is, for simplicity, restricted to first
derivatives. In the future, the interface could be easily expanded if
specific use cases require higher order derivatives.

In scientific computing, linear functions such as, e.g., spherical
harmonic transforms are widespread. If the function \(f\) is linear,
differentiation becomes trivial. Specifically for a linear function
\(f\), the pushforward respectively the \texttt{jvp} of \(f\) is
identical to \(f\) itself and independent of the position at which it is
computed. Expressed in formulas, \(\partial f(x)(dx) = f(dx)\) if \(f\)
is linear in \(x\). Analogously, the pullback respectively the
\texttt{vjp} becomes independent of the initial position and is given by
the linear transpose of \(f\), thus
\((\partial f(x))^{T}(dy) = f^T(dy)\). Also, all higher order
derivatives can be expressed in terms of \(f\) and its transpose. To
make use of these simplifications, \texttt{JAXbind} provides a special
interface for linear functions, supporting higher order derivatives,
only requiring an implementation of the function and its transpose.

@@

Currently, \texttt{JAXbind} only has CPU but no GPU support. With some
expertise on Python bindings for GPU kernels adding GPU support should
be fairly simple. We especially want to highlight that the interfacing
with the JAX automatic differentiation engine is identical for CPU and
GPU.
